The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-02-27 src/ms.tex
Start Table Mixin Columns
Signed-off-by: nstarman <nstarkman@protonmail.com>

% Links to documentation: subpackages
\newcommand{\astropysubpkg}[1]{\href{http://docs.astropy.org/en/stable/#1/index.html}{\texttt{astropy.#1}}\xspace}
\newcommand{\astropyiosubpkg}[1]{\href{http://docs.astropy.org/en/stable/io/#1/index.html}{\texttt{astropy.io.#1}}\xspace}
\newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}
\newcommand{\astropycosmology}{\astropysubpkg{cosmology}}
\newcommand{\astropycosmologyunits}{\href{https://docs.astropy.org/en/stable/cosmology/units.html}{\texttt{astropy.cosmology.units}}}
\newcommand{\astropyunits}{\astropysubpkg{units}}
\newcommand{\astropycoordinates}{\astropysubpkg{coordinates}}
\newcommand{\astropyconstants}{\astropysubpkg{constants}}
\newcommand{\astropytime}{\astropysubpkg{time}}
\newcommand{\astropymodeling}{\astropysubpkg{modeling}}

@@

\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
GalacticLSR}{GalacticLSR}}
\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
\newcommand{\astropyModel}{\astropyapidoc{modeling.Model}{Model}}
\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
\newcommand{\astropyMasked}{\astropyapidoc{utils.masked.Masked}{Masked}}
\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}

@@

  \paragraph{Astropy Fits in Time}    % IN PROGRESS (ADAPTING FROM THE WHATSNEW).

    As anyone who has worked with FITS files knows, the FITS standard is both
    detailed and very flexible. While this means FITS is useful across many
    domains, working with the files can be involved. Expanding on the WCS and
    related work, the \astropyfits package now provides support for reading and
    writing native \astropyTime table columns using the FITS Time standard
    \citep{FITS-Time:2015}. For backward compatibility, this feature may be
    turned off. Implementation of this feature was made possible through the
    support of the Google Summer of Code 2017 program.
    % TODO: is there a way to cite this
    To learn more about this project please see the final report,
    \href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
    mixin protocol for seamless interoperability}.

@@

    Numerous \python libraries have custom table implementations
    \begin{itemize}
      \item (Support use of SkyCoord as Column happened before) and in table vstack, dstack, and insert\_row
      \item Converting SkyCoord to QTable (1 sentence addition to above)
      \item Added support for \package{dask} arrays in tables
      \item Added support for registering array-like objects as mixin columns
    \end{itemize}

@@

  % MHvK: maybe do this before the structure? Change in structure
  % partially driven by this?
    The improvements to \astropy's I/O architecture are mirrored in the increased support for I/O options.

    \begin{itemize}
      \item ASDF : JSWST  and models can be stored with ASDF.
      \item ECSV : Support for multidimensional and object columns in ECSV
      \item FITS : Support for writing Dask arrays to FITS files, and Append table to existing FITS file
      \item Support for reading, writing, and converting Cosmology
      \item MRT Support for reading and writing tables to MRT format : GSOC
      \item Support for reading and writing tables to Parquet format
      \item Support for reading and writing tables to QDP format
      \item ASDF read write
    \end{itemize}

@@

    \astropyiosubpkg{registry} is a powerful way to define input and output
    (I/O) functions for \astropypkg objects, such as reading or writing to a
    file. \astropy uses this internally for the I/O methods in \astropyTable and
    \astropyCosmology (I/O added v5.0+), and users can register custom I/O to
    extend the options on these classes.

>>>>

% Links to documentation: subpackages
\newcommand{\astropysubpkg}[1]{\href{http://docs.astropy.org/en/stable/#1/index.html}{\texttt{astropy.#1}}\xspace}
\newcommand{\astropyiosubpkg}[1]{\href{http://docs.astropy.org/en/stable/io/#1/index.html}{\texttt{astropy.io.#1}}\xspace}
\newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}
\newcommand{\astropycosmology}{\astropysubpkg{cosmology}}
\newcommand{\astropycosmologyunits}{\href{https://docs.astropy.org/en/stable/cosmology/units.html}{\texttt{astropy.cosmology.units}}}
\newcommand{\astropyunits}{\astropysubpkg{units}}
\newcommand{\astropycoordinates}{\astropysubpkg{coordinates}}
\newcommand{\astropyconstants}{\astropysubpkg{constants}}
\newcommand{\astropytable}{\astropysubpkg{table}}
\newcommand{\astropytime}{\astropysubpkg{time}}
\newcommand{\astropymodeling}{\astropysubpkg{modeling}}
\newcommand{\astropywcs}{\astropysubpkg{wcs}}
\newcommand{\astropyfits}{\astropyiosubpkg{fits}}

@@

\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
GalacticLSR}{GalacticLSR}}
\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
\newcommand{\astropyFitsColumn}{\astropyapidoc{io.fits.Column}{Column}}
\newcommand{\astropyModel}{\astropyapidoc{modeling.Model}{Model}}
\newcommand{\astropyTableColumn}{\astropyapidoc{table.Column}{Column}}
\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
\newcommand{\astropyMasked}{\astropyapidoc{utils.masked.Masked}{Masked}}
\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}

@@

@@

    % TODO: this is very rough. Probably needs a bit of a re-write.

    Within \astropypkg and \python there are numerous ways to represent and
    store array-valued data. Some of the differences are historical:
    \astropyapidoc{table.Column}{table.Column} and
    \astropyapidoc{io.fits.Column}{io.fits.Column} are not the same. Some
    differences are computational: \package{numpy}, \package{cupy}, and
    \package{dask} arrays are almost API identical, but are optimized for
    different compute needs. Lastly, some differences are inherent:
    \astropyQuantity, \astropyTime and \astropySkyCoord represent fundamentally
    different types of objects.

    \astropy believes however array-valued data is stored, it should be usable
    as a column in a table: historical differences are baked into stable API,
    computational needs can differ, and some objects have extra information. The
    \astropytable and \astropyfits table implementations allow for mixin-columns
    (in general a mixin column may contain multiple data components as well as
    object attributes beyond the standard column). A well defined protocol has
    been developed to allow the usage of these mixin columns while allowing the
    object to ``round-trip'' through tables with no loss of data or attributes.
    With this protocol, it is now possible to store not only standard
    \astropyTableColumn objects to a FITS table
    \href{https://docs.astropy.org/en/stable/io/fits/api/hdus.html}{HDU}, but
    also any \astropy native objects within a \astropyTable. This includes
    \astropyTime, \astropyQuantity, \astropySkyCoord, and many others. For
    objects not already covered by the mixin protocol, functions can be
    registered with \astropyTable to convert, on-the-fly, any array-like object
    into a `mixin' column that can be used in a table. As an example, this is
    used to provide the seamless integration of \package{dask} arrays in
    \astropyTable, allowing computer cluster-scale data to be stored in a
    \astropyTable.

  \paragraph{Astropy Fits in Time}
    % TODO! maybe merge with previous section

    Time as a dimension in astronomical data presents challenges in its
    representation in FITS files. The standard has therefore been extended to
    describe rigorously the time coordinate in the World Coordinate System (WCS)
    framework \citep{FITS-Time:2015}.

    \astropyfits can store this data in a standard \astropyFitsColumn (for
    details, see the
    \href{https://docs.astropy.org/en/stable/io/fits/index.html}{FITS
    documentation}), but will not ``understand'' the metadata and attributes.
    Parsing the column to a useful time representation would be done manually.
    Thankfully, through the support of the Google Summer of Code 2017 program,
    % TODO: is there a way to cite this?
    the \astropyfits package now allows \astropyTime as a `mixin' column.
    \astropyTime columns in FITS tables ``understand'' the metadata -- scale,
    reference, position, offset, etc. -- while also providing a similar
    interface as the regular \astropyFitsColumn objects. For backward
    compatibility, this feature may be turned off. To learn more about this
    project please see the final report,
    \href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
    mixin protocol for seamless interoperability}.

@@

    % TODO! need good transition sentence. This one is bad.
    The column mixin options in tables are mirrored in the increased support for
    I/O options.

    \begin{itemize}
      \item ECSV : Support for multidimensional and object columns in ECSV
      \item FITS : Support for writing Dask arrays to FITS files, and Append table to existing FITS file
      \item MRT Support for reading and writing tables to MRT format : GSOC
      \item Support for reading and writing tables to Parquet format
      \item Support for reading and writing tables to QDP format
      \item ASDF read write
    \end{itemize}

    Also stuff that isn't a table

    \begin{itemize}
      \item Support for reading, writing, and converting Cosmology
    \end{itemize}

@@

    % TODO! better transition
    \astropyiosubpkg{registry} is a powerful way to define input and output
    (I/O) functions for \astropypkg objects, such as reading or writing to a
    file. \astropy uses this internally for the I/O methods in \astropyTable and
    \astropyCosmology (I/O added v5.0+), and users can register custom I/O to
    extend the options on these classes.
